\textsf{Are large language models superhuman chemists?}

### 2024-10-28 src/tex/appendix.tex
Update appendix.tex
feat: add comments about no data from benchs (#80)
Co-authored-by: Kevin M Jablonka <32935233+kjappelbaum@users.noreply.github.com>

@@

In addition, as shown in \Cref{tab:chembench_corpus_cognitive} the \chembench corpus can be divided considering the different skills needed to solve the questions.

\begin{xltabular}{\textwidth}{p{2.7cm}p{4em}X}
        \caption{\textbf{Examples for each of the required skills considered in the \chembench corpus.} The table shows the number of questions for each skill and an example question.} \\

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 919 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 987 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By atomic force microscopy \\
        & & B. By transform infrared spectroscopy \\
        & & C. By measuring a photoluminescence spectrum \\
        & & D. By absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 128 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

    \label{tab:chembench_corpus_cognitive}
\end{xltabular}

@@

@@

% In addition, we analyzed the performance on questions which requires calculation. For this, we manually labeled questions that require multiple calculation steps.
% We find that the ranking of models is different for questions with our without calculations (\Cref{fig:requires_cal}).

% \begin{figure}
%     \centering
%     \includegraphics{figures/model_overall_cal.pdf}
%     \caption{\textbf{Overall model performance for questions with and without calculation steps.} We find that the ranking of models changes if we evaluate them on questions that require and do not require calculations, respectively.}
%     \script{requires_calculation_plot.py}
%     \label{fig:requires_cal}
% \end{figure}

@@

\clearpage
\subsection{Summary of trends and recommendations}

\oone consistently leads across both \ChemBench corpus and \chembenchmini.  \ClaudeThreeFiveSonnet, \GPTFourO, \LlamaThreeOneFourZeroFiveBInstruct form the next tier of strong performers. These models show robust performance across all skills and difficulty levels.

Larger models generally perform better (\LlamaThreeOneFourZeroFiveBInstruct > \LlamaThreeOneSeventyBInstruct > \LlamaThreeOneEightBInstruct). However, smaller but well-designed models can compete with larger models. For example, \GemmaTwoNineBIt has an overall accuracy of 0.48 on \ChemBench corpus, which is only 10\% less accurate than the much larger \LlamaThreeOneFourZeroFiveBInstruct model.
One can observe a clear progression of performance across model families (e.g., \ClaudeThreeFiveSonnet >  \ClaudeThree > \ClaudeTwo or \GPTFourO > \GPTFour > \GPTThreeFiveTurboZeroT). Newer versions consistently outperform their predecessors, indicating improvements in scientific understanding of these model families.
Scores on knowledge-intensive tasks are typically lower than those on calculation and reasoning-intensive questions. Performance on reasoning-intensive questions in \chembench tends to be the highest.

Based on these findings, some scope for improving these models could be to focus on enhancing knowledge-based training, possibly through improved pre-training on chemistry-specific texts or with chemistry knowledge-base integration. 
Models show the strongest performance in reasoning tasks on the \chembench corpus, which suggests that current training approaches are good at developing logical/reasoning capabilities for basic problem-solving of university-level exams. 
However, performance drops significantly for advanced tasks, suggesting that more advanced chemistry problems with complex, multi-step solutions should be included in training or finetuning.

>>>>

\subsubsection{Curation workflow}
Questions were added via pull requests to the \chembench repository on GitHub.
This allowed for a manual review of each question by expert reviewers (with backgrounds in chemistry, materials science, chemical engineering, and physics).
The reviews were conducted directly on the GitHub platform, where our entire curation history is also publicly available.

The general guidelines followed by the reviewer are the following:

\begin{itemize}
    \item \textbf{Originality:} Questions should not be readily findable online or in other easily accessible sources (example \url{https://github.com/lamalab-org/chem-bench/pull/392#discussion_r1694299474})
    \item \textbf{Ambiguity:} Questions with unclear wording or multiple interpretations (example \url{https://github.com/lamalab-org/chem-bench/pull/420#discussion_r1698147159} )
    \item \textbf{Factual or heuristic Errors: }Questions containing factual inaccuracies or misconceptions are not included (example \url{https://github.com/lamalab-org/chem-bench/pull/389#discussion_r1686187301})
    \item \textbf{Out of Scope:} Questions outside the realm of chemistry are rejected.
    \item \textbf{Clarity and Difficulty: } They should pose a challenge and encourage exploration within the chemical domain.
    \item \textbf{Contribution to Dataset Diversity: }Questions should cover a wide range of chemical concepts and sub-disciplines. They should add value by expanding the breadth of the dataset. That is, questions already multiple (>10) times in the corpus in a similar form are rejected.
\end{itemize}

Reviewers also solved the questions to verify the answers. They also performed web searches to ensure questions were not easily found online. The reviewers often guided the revision process to ensure the question aligned with the guidelines. Questions that don't meet the criteria are either rejected or suggested for revision and most often, they are modified to a new question. Reviewers also provide feedback on the skill and difficulty annotations.

In addition to the manual review, we also performed automated checks to ensure the quality of the questions. The schemas, Latex templating, and other formatting aspects are checked automatically using GitHub Actions.

While adding questions from existing benchmarks might seem to be another good source of semi-automatically generated data, we prioritized the diversity of the data and avoided data contamination while addressing the guidelines above and in \Cref{sec:desired-properties}.
However, even though we decided not to include questions from other previously published chemistry-focused benchmarks into the \chembench corpus, the framework is flexible enough to be readily extended with questions from other benchmarks.

\subsubsection{Composition}

@@

In addition, as shown in \Cref{tab:cb_skillset} the \chembench corpus can be divided considering the different skills needed to solve the questions. The plot shows a balanced distribution of the required skills in the \chembench corpus. \Cref{tab:cb_skillset} shows an example question for each skill category.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/CB_SKILL_FIG_V1.pdf}
    \caption{\textbf{Composition of the required skills considered in the \chembench corpus.} The circular plot shows the distribution of required skills in the \chembench corpus.}
    \label{fig:cb_skillset}
\end{figure}

\begin{table}
    \centering
    \small
    \caption{\textbf{Examples for each of the required skills considered in the \chembench corpus.} The table shows the number of questions for each skill and an example question. Note that the total count in this table is bigger than the \chembench corpus. This is because the same question can be annotated with two different \enquote{requires} skills, i.e. Reasoning and Calculation.} \\
    \label{tab:chembench_corpus_cognitive}
    {\fontsize{8}{9}\selectfont
        \begin{tabularx}{\textwidth}{X}
            \toprule
            \multicolumn{1}{c}{\textbf{Knowledge} 919 Questions} \\
            \midrule
            Which of the following salts is an acidic salt? \\
            A. \ce{NH4Cl} \\
            B. \ce{Na2CO3} \\
            C. \ce{NaH2PO4} \\
            D. \ce{Zn(OH)Cl} \\
            \midrule
            \multicolumn{1}{c}{\textbf{Reasoning} 987 Questions} \\
            \midrule
            How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
            A. By atomic force microscopy \\
            B. By transform infrared spectroscopy \\
            C. By measuring a photoluminescence spectrum \\
            D. By absorption spectroscopy \\
            \midrule
            \multicolumn{1}{c}{\textbf{Intuition} 1001 Questions} \\
            \midrule
            Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
            A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
            B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
            \midrule
            \multicolumn{1}{c}{\textbf{Calculation} 128 Questions} \\
            \midrule
            Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
            \bottomrule
        \end{tabularx}
    }
\end{table}

In analyzing the performance of state-of-the-art models on the ChemBench corpus, we identified several chemical topics where the models consistently struggled to provide correct answers. The following table \cref{tab:chembench_corpus_models_failed_topic} presents examples of challenging questions across different chemical domains, highlighting specific areas where current models need improvement

\begin{table}
    \centering
    \small
    \caption{\textbf{Examples of topics evaluated in the \chembench corpus that the state-of-the-art model failed to answer correctly.} The table displays the percentage of questions in the corpus for each topic, along with example questions and the percentage of questions that were not answered correctly by the models.} \\
    \label{tab:chembench_corpus_models_failed_topic}
    {\fontsize{8}{9}\selectfont
        \begin{tabularx}{\textwidth}{X}
            \multicolumn{1}{c}{\textbf{Analytical Questions} 166 Questions (5.8\%) (26.5\%)} \\
            Which of the following methods can be used to support the structure elucidation of planar molecules by examining the topography of individual molecules? \\
            A. DESI-mass spectrometry \\
            B. XPS \\
            C. atomic force microscopy \\
            D. scanning tunneling microscopy \\
            E. secondary ion mass spectrometry \\
            \multicolumn{1}{c}{\textbf{Chemical Preference} 1001 Questions (35\%) (8.5\%)} \\
            Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
            [START\_SMILES]N\#Cc1ccc(OCCCN2CC3CN\-(CCNS(=O)(=O)c4ccc(F)cc4)CC(C2)O3)cc1\-[END\-\_SMILES] \\
            [START\_SMILES]O=C1CC(c2ccc(CC(NS(=O)\-(=O)\-c3cc(Cl)cc(Cl)c3)c3nc4ccccc4[nH]3)cc2)S\-(=O)\-(=O)N1[END\_SMILES] \\
            \multicolumn{1}{c}{\textbf{General Chemistry} 152 Questions (5.3\%) (3.2\%)} \\
            Why \ce{SCl6} does not exist if other similar molecules such as \ce{SF6} exist? \\
            A. \ce{SCl6} does not exist because of steric effects \\
            B. \ce{SCl6} does not exist because of the high electronegativity of chlorine \\
            C. \ce{SCl6} does not exist because of the high ionization potential of the chlorine \\
            D. \ce{SF6} exist because of the high electronegativity of fluorine \\
            \multicolumn{1}{c}{\textbf{Inorganic Chemistry} 94 Questions (3.3\%) (4.2\%)} \\
            What is the spin-only magnetic moment (in units of Bohr-magnetons) for the low-spin case of the metal complex \ce{RhCl(CO)(PPh3)2}?\\
            \multicolumn{1}{c}{\textbf{Materials Science} 89 Questions (3.1\%) (0.9\%)} \\
            What are the conditions for the observation of effects of quantum confinement in quantum dots? \\
            A. All dimension of the material must be smaller than exciton Bohr radius \\
            B. One dimension of the quantum dot must be smaller than exciton Bohr radius \\
            C. The movement od charge carriers is limited in all spatial dimensions \\
            \multicolumn{1}{c}{\textbf{Organic Chemistry} 441 Questions (15.5\%) (0.5\%)} \\
            What makes acrylates more reactive than typical hydrocarbons? \\
            A. The carbonyl group \\
            B. The double bond \\
            C. The hydroxyl group \\
            D. The vinyl group \\
            \multicolumn{1}{c}{\textbf{Physical Chemistry} 179 Questions (6.3\%) (13.4\%)} \\
            Why are polar molecules more soluble in water than non-polar molecules? \\
            A. Non-polar molecules should break hydrogen bonds in water to dissolve \\
            B. The entropy of the system increases when non-polar molecules dissolve in water \\
            C. Water has a dipole and can easier isolate polar molecules from each other \\
            D. Water makes stable dipole-dipole interactions with polar molecules \\
            \multicolumn{1}{c}{\textbf{Technical Chemistry} 42 Questions (1.5\%) (4.7\%)} \\
            Which of the following statements is/are true about granulation shaping of solid catalysts? \\
            A. Broader product diameter distribution than pellets or extrudates \\
            B. Facilitates reactor upscaling \\
            C. It is also called snow-ball \\
            D. Not suitable to prepare egg-shell catalysts \\
            \multicolumn{1}{c}{\textbf{Toxicity/Safety} 690 Questions (24.2\%) (40.4\%)} \\
            What determines the flammability of polyols? \\
            A. Their chemical structure \\
            B. Their low volatility \\
            C. Their odor threshold \\
            D. Their reaction with acids \\
        \end{tabularx}
    }
\end{table}

To better understand the cognitive demands of chemical assessment, we categorized the questions in the ChemBench corpus according to the primary skills required to solve them. The table \Cref{tab:chembench_corpus_models_failed_cognitive} below shows representative examples of questions that state-of-the-art models failed to answer correctly across different cognitive skill categories, including knowledge, reasoning and intuition.
\begin{table}
    \centering
    \small
    \caption{\textbf{Examples of required skills evaluated in the \chembench corpus that the state-of-the-art model failed to answer correctly} The table displays the number of questions for each skill, along with an example question that the model failed to answer.} \\
    \label{tab:chembench_corpus_models_failed_cognitive}
    {\fontsize{8}{9}\selectfont
        \begin{tabularx}{\textwidth}{X}
            \toprule
            \multicolumn{1}{c}{\textbf{Knowledge} 919 Questions} \\
            \midrule
            Which of the following are true for phosphoric acid? \\
            A. Is a weak acid, so it is not toxic \\
            B. Should never be tasted \\
            C. Though a component of soft drinks, reacts vigorously with water when in concentrated form \\
            \midrule
            \multicolumn{1}{c}{\textbf{Reasoning} 987 Questions} \\
            \midrule
            Which of the following elements has the largest atomic radius? \\
            A. \ce{Fr} \\
            B. \ce{He} \\
            C. \ce{Li} \\
            D. \ce{Rb} \\
            \midrule
            \multicolumn{1}{c}{\textbf{Intuition} 1001 Questions} \\
            \midrule
            Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
            A. [START_SMILES]CCc1c(CCOC)nn(-c2ccccc2)c1-c1ccccc1[END_SMILES] \\
            B. [START_SMILES]Cc1cc(/C=C2\\C(=N)N3N=C(S(C)(=O)=O)SC3=NC2=O)c(C)n1Cc1ccccc1[END_SMILES] \\
            \midrule
            \multicolumn{1}{c}{\textbf{Calculation} 128 Questions} \\
            \midrule
            The cell potential for the overall cell reaction during electrolysis of water, \ce{2H2O(l) -> 2H2(g) + O2(g)} is \SI{-1.23 V}. What is the standard electrode potential for the half reaction, \ce{2H2O(l) -> O2(g) + 4H+(aq) + 4e-} in V? \\
            \bottomrule
        \end{tabularx}
    }
\end{table}

@@

\Cref{fig:performance_corpus_and_tiny} shows the performance of the models on the \chembench corpus and the \chembenchmini subset. The relative performance difference between both is quite similar across most models. 
This makes \chembenchmini subset a reliable subset for human baseline comparison and particularly valuable for rapid prototyping and initial model assessment phases.

An interesting observation is the significant impact of chemical preference tasks on \GPTFour's scores. A detailed breakdown of overall accuracy into scores on different skills and difficulty levels is provided in \Cref{table:performance_table} and \Cref{tab:performance_table_human_subset} for \chembench corpus and the \chembenchmini subset, respectively.

@@

\begin{figure}[htb]
    \centering
    \includegraphics{figures/corpus_human_comparison.pdf}
    \caption{\textbf{
        Performance of the models on the \chembench corpus and \chembenchmini subset.} The bar plot shows the fraction of questions that were answered completely correctly, highlighting the relative performance of different models on both the \chembench corpus and the \chembenchmini subset.
        We see that the model ranking remains fairly consistent across both sets}  
    \label{fig:performance_corpus_and_tiny}
    \script{corpus_humanset_performance.py}
\end{figure}

@@

### 2024-10-28 src/tex/ms.tex
feat: add comments about no data from benchs (#80)
Co-authored-by: Kevin M Jablonka <32935233+kjappelbaum@users.noreply.github.com>

\begin{table}[!h]
    \centering
    \caption{\textbf{Overview of sources of the curated questions}. The table provides an overview of the types of sources the questions have been curated from. Detailed sources are available in the source data on GitHub. Questions without source have been curated completely from scratch. Questions based on lecture notes or URL have been curated based on content presented in those resources. All questions have been rephrased, annotated, and reviewed before being added to the corpus.}
    \variable{output/sources_table.tex}
    \label{tab:sources}
\end{table}

>>>>

\begin{table}[!h]
    \centering
    \caption{\textbf{Overview of sources of the curated questions}. The table provides an overview of the types of sources the questions have been curated from. Detailed sources are available in the source data on GitHub. Questions without source have been curated completely from scratch. Questions based on lecture notes or URL have been curated based on content presented in those resources. All questions have been rephrased, annotated, and reviewed before being added to the corpus. Note that the chemical preference questions are the main corpus of the semi-automatically generated group with \variable{output/question_count_per_dir/json_file_counts_preference.txt} questions of the total \variable{output/automatically_generated.txt}.}
    \variable{output/sources_table.tex}
    \label{tab:sources}
\end{table}
